### 2015-01-16 cav-2015-tex/evaluation.tex
Added the technnical stuff

The results of the experiments are presented in Table
\ref{tab:results}. Each row contains the results for the specified $k$-bound 
for each artifact evaluated. The columns show the total
wall clock time in seconds, states explored, and paths generated for
each of the algorithms. The headings \gsetxt{}, L\#, and SL correspond
to the Generalized Symbolic Execution, lazier\#, and \symtxt{}
algorithms, respectively. Note that we do no present the results for
the lazier algorithm since they are almost identical to that of lazier\#
for our examples. Table entries containing '-' correspond to test runs
that exceeded the allotted time bound of 10 minutes.

In most tests, run times strongly favor the \symtxt{} algorithm, especially for
large $k$-values. Performance improvement ranges from $4.8x$ for
BinarySearchTree at $k$=3, to $118x$ for LinkedList at $k$=6. In fact,
for certain $k$-bounds a number of experiments completed exploration
using the \symtxt{} algorithm in a few seconds whereas \gsetxt{} or lazier\# were
unable to finish, e.g., BinarySearchTree for $k$=4 and LinkedList for
$k$=8. These results demonstrate the potential of using the \symtxt{} algorithm 
to handle data input for complex types. The lower running times
show that we can prove properties using the \symtxt{} algorithm faster than
\gsetxt{}.

% Some of this discrepancy may be accounted for by the fact that
%\symtxt{} is equipped for incremental solving, however, during
%informal testing with the incremental solver turned off, \symtxt{}
%usually performed better than \gsetxt{} and lazier\#, especially in
%test runs with high k-bounds.

@@

The number of states varies between algorithms, for example, \gsetxt{}
has additional points of nondeterminism during field reads, but in
contrast reference compares are completely deterministic. Thus, in
example programs with large numbers of reference compares, such as
TreeMap, state counts for lazier\# and the \symtxt{} algorithm may
exceed those for \gsetxt{}. We observed that the additional states
generated by the summary heap algorithm are unsatisfiable at the point
of reference compares; this is why they do not contribute any
additional branches in the final symbolic execution tree. Furthermore,
the larger state count does not impact the overall runtime of the
technique.

@@

In summary, the benefits of avoiding
initialization-based non-determinism outweigh the increased
complexity in the constraints due to the advances made in SMT
solvers in these examples. In fact, the \symtxt{} algorithm can analyze certain types of programs with
orders of magnitude greater efficiency than with \gsetxt{} or
lazier\#, while covering exactly the same control flow paths.

>>>>

The results of the experiments are presented in Table
\ref{tab:results}. Each row contains the results for the specified $k$-bound 
for each artifact evaluated. The columns show the total
wall clock time in seconds, states explored, and paths generated for
each of the algorithms. The headings \gsetxt{}, L\#, and SL correspond
to the Generalized Symbolic Execution, Lazier\#, and \symtxt{}
algorithms, respectively. Note that we do no present the results for
the Lazier algorithm since they are almost identical to that of Lazier\#
for our examples. Table entries containing '-' correspond to test runs
that exceeded the allotted time bound of 10 minutes.

In most tests, run times strongly favor the \symtxt{} algorithm, especially for
large $k$-values. Performance improvement ranges from $4.8x$ for
BinarySearchTree at $k$=3, to $118x$ for LinkedList at $k$=6. In fact,
for certain $k$-bounds a number of experiments completed exploration
using the \symtxt{} algorithm in a few seconds whereas \gsetxt{} or Lazier\# were
unable to finish, e.g., BinarySearchTree for $k$=4 and LinkedList for
$k$=8. These results demonstrate the potential of using the \symtxt{} algorithm 
to handle data input for complex types. The lower running times
show that we can prove properties using the \symtxt{} algorithm faster than
\gsetxt{}.

% Some of this discrepancy may be accounted for by the fact that
%\symtxt{} is equipped for incremental solving, however, during
%informal testing with the incremental solver turned off, \symtxt{}
%usually performed better than \gsetxt{} and Lazier\#, especially in
%test runs with high k-bounds.

@@

The number of states varies between algorithms, for example, \gsetxt{}
has additional points of nondeterminism during field reads, but in
contrast reference compares are completely deterministic. Thus, in
example programs with large numbers of reference compares, such as
TreeMap, state counts for Lazier\# and the \symtxt{} algorithm may
exceed those for \gsetxt{}. We observed that the additional states
generated by the summary heap algorithm are unsatisfiable at the point
of reference compares; this is why they do not contribute any
additional branches in the final symbolic execution tree. Furthermore,
the larger state count does not impact the overall runtime of the
technique.

@@

In summary, the benefits of avoiding
initialization-based non-determinism outweigh the increased
complexity in the constraints due to the advances made in SMT
solvers in these examples. In fact, the \symtxt{} algorithm can analyze certain types of programs with
orders of magnitude greater efficiency than with \gsetxt{} or
Lazier\#, while covering exactly the same control flow paths.

### 2015-01-16 cav-2015-tex/summary-heap.tex
Added the technnical stuff

@@

>>>>

\section{Summary Heap Framework}
The summary heap algorithm is presented using Javalite, an
executable formalization of a Java-like language~\cite{saints-MS}. Javalite retains 
the essence of Java and other object-oriented languages but with formalizations 
in both PLT Redex and Coq so that the outcome of any program has a precise
and deterministic meaning.

Javalite is a syntactic machine defined as rewrites on a string.  The
semantics are small-step using the CESK architecture with a (C)ontrol string
representing the expression being evaluated, an (E)nvironment for
local variables, a (S)tore for the heap, and a (K)ontinuation for what
is to be executed next.  In this work we change the semantics of the original
 Javalite machine and its machine syntax to represent (i) GSE~\cite{GSE03}, and (ii) the summary heap algorithm
presented in this work.

The surface syntax (input) for Javalite is shown 
in~\figref{fig:surface-syntax}, and the machine syntax (state)
in~\figref{fig:machine-syntax}. Terminals are in bold face while
non-terminals are italicized. Ellipses indicate zero or more
repetitions. Tuples omit the commas for compactness. The language
itself uses s-expressions for convenience.

\input{surface-syntax}
\input{machine-syntax}

A Javalite program, $\cfgnt{P}$, is a registry of classes, $\mu$, with
a tuple indicating a class, $\cfgnt{C}$, and a method, $\cfgnt{m}$,
where execution starts. The only primitive type in Javalite is
Boolean. Classes have non-primitive fields, $([\cfgnt{C}\ \cfgnt{f}]\ ...)$, and
methods, $(\cfgnt{M}\ ...)$.\footnote{Summary heaps are able to support non-primitive fields (see \secref{sec:eval}). They are omitted to focus and simplify the presentation.} Methods are limited to a single
parameter in our Javalite machine. Expressions, $\cfgnt{e}$,  
include statements, and they
 use '$\cfgt{:=}$' to indicate assignment and '$\cfgt{=}$' to indicate comparison.
The dot-operator for field access is replaced by '$\cfgt{\$}$', and the dot-operator
for method invocation is replaced by '$\cfgt{@}$' since '.' is reserved in PLT Redex. There is no explicit return
statement in Javalite; rather, the value of the last expression is
used as the return value. A variable is always indicated by \cfgnt{x}
and a value by \cfgnt{v}. A value can be a reference in the heap, $\cfgnt{r}$, or any of the special values shown in~\figref{fig:surface-syntax}. 
We assume that only
type correct programs are given as input to the machine.

The state of the Javalite machine (\figref{fig:machine-syntax}),
$\cfgnt{s}$, includes the program registry, $\mu$, a pair of functions defining the heap
$\lp\cfgnt{L}\ \cfgnt{R}\rp$, a constraint,
$\phi$, defined over references in the heap, the environment, $\eta$,
for local variables, and the continuation $\cfgnt{k}$.\footnote{The registry $\mu$ never changes so it is omitted from the state tuple in the rest of the presentation for compactness.}  The more
complex structures, such as the environment and heap, are defined as lists that
start with empty, \cfgnt{mt}. The rewrite rules that define the
semantics treat these lists as partial functions. As such,
$\eta(\cfgnt{x}) = \cfgnt{r}$ is the reference $\cfgnt{r}$ associated with variable
$\cfgnt{x}$. The notation $\eta^\prime = \eta[\cfgnt{x} \mapsto
  \cfgnt{v}]$ defines a new partial function $\eta^\prime$ that is
the same as $\eta$ except that the variable $\cfgnt{x}$ now maps to
$\cfgnt{v}$.

The continuation $\cfgnt{k}$ helps accomplish the small-step semantics
by indicating where the hole, $\cfgt{*}$, is located, by storing
temporary computations, and by keeping track of the next
computation. For example, the continuation
$\lp\cfgt{*}\ \cfgt{\$}\ \cfgnt{f} \rightarrow \cfgnt{k}\rp$ indicates
that the machine is currently computing the expression for the object
reference on which the field $\cfgnt{f}$ is going to be accessed. Once
the field access is complete, the machine continues with $\cfgnt{k}$.

@@

\subsection{Generating Summary Heaps}
\label{sec:precise}
The summary
heap algorithm only
partitions the heap at points of divergence in the control flow graph
such as field accesses or writes (execeptional control flow due to null
pointers) or comparisons of references (program control flow). There
are three sets of rewrite rules specific to the summary heap
algorithm: (i) rules to initialize symbolic references, (ii) rules to
complete field access and field write, and (iii) rules to check
equality and inequality of references. The other rules are common to
both $\rsym$ and $\rgse$. The complete set of rules are shown in the
supplemental attached with the paper.

Initialization of references in GSE ($\rgse$ relation) is a
non-deterministic choice where the reference in the heap is
initialized to null, a new instance of the object, or to any type
compatible object in the input heap (e.g., previously initialized by
$\rgse$). In contrast, initialization in the $\rsym$ relation creates
a single summary heap containing all of the possible initialization
choices.

\input{summarize-precise-incomplete}

\begin{figure*}
\begin{tabular}[c]{l}
\scalebox{1.0}{\usebox{\boxPI}} \\
\end{tabular}
\caption{The summary machine, $s ::= \lp\cfgnt{L}\ \cfgnt{R}\ \cfgnt{r}\ \cfgnt{f}\ \cfgnt{C}\rp$, with $s\rsum^*s^\prime =  s \rsum \cdots \rsum s^\prime \rsum s^\prime$.}
\label{fig:symInit}
\end{figure*}

The initialization rules, \figref{fig:symInit}, are invoked whenever a
field pointing to an uninitialized reference is accessed. The
interaction with the solver in the definition of the rules is denoted
by $\mathbb{S}(\phi)$ which returns true if $\phi$ is satisfiable. The
summarize-end rule becomes active when there is nothing to
initialize. This condition is determined by the function
$\mathbb{UN}(\cfgnt{L}, \cfgnt{R}, \cfgnt{r}, \cfgnt{f}) =
\{\lp\phi\ \cfgnt{l}\rp\ ...\}$ which returns constraint-location
pairs where the field $\cfgnt{f}$ is uninitialized.
\[
\begin{array}{rcl}
\mathbb{UN}(\cfgnt{L}, \cfgnt{R}, \cfgnt{r}, \cfgnt{f}) & = &\{ \lp\phi\ \cfgnt{l}\rp \mid \lp \phi\ \cfgnt{l}\rp  \in \cfgnt{L}\lp \cfgnt{r}\rp  \wedge \\
& & \ \ \ \ \exists \phi^\prime \lp \lp \phi^\prime\ \cfgnt{l}_\mathit{un}\rp  \in \cfgnt{L}\lp \cfgnt{R}\lp l,\cfgnt{f}\rp\rp \wedge \\
& & \ \ \ \ \ \ \ \ \mathbb{S}\lp \phi \wedge \phi^\prime \rp\rp\}\\
\end{array}
\]
A pair in the set means the field in the location maps to a reference
that then points to $\cfgnt{l}_\mathit{un}$, and that the path to that
location is feasible on some aliasing assignment.

% The cardinality of the set, $\theta$ is never
%greater than one in GSE and the constraint is always satisfiable
%because all constraints are constant. This property is relaxed in GSE
%with heap summaries.

In~\figref{fig:symInit}, given the uninitialized set $\Lambda$ for
field $f$, the $\mathit{min}_l$ function returns
$(\phi_x\ \cfgnt{l}_x)$ based on a lexicographical ordering of
uninitialized locations in $\Lambda$ to make initialization
deterministic (i.e., the same heap is always initialized in the same
way as order determines how many aliases are available in the input heap at any given point in initialization). 
%The $\mathrm{fresh}_\cfgnt{l}(\cfgnt{C})$ call returns a new location of type $C$. 
The set $\rho$ contains the set of
reference-location pairs from the input heap that are potential
aliases. Minimizing over the set of references in the pre-image ensures
that any pair in $\rho$ is the reference $\cfgnt{r}_a$ and the location $\cfgnt{l}_a$ materialized at the same time in some prior initialization (i.e., they were once
$\cfgnt{r}_f$ and $\cfgnt{l}_f$). Aliasing constraints
in the new summary heap reason over these original creators; otherwise aliasing and local value set
computation breaks down.  The reference partitions with their
monotonicity make this determination possible.

There are three sets corresponding to the non-deterministic choices in
GSE with a fourth that preserves the pre-initialized heap structure.
Note that the constraints always preserve the access path to the location which is $\phi_x$.
(i) $\theta_\mathit{null}$ asserts the condition under which
$\cfgnt{l}_\mathit{null}$ is possible (any time the solver assigns
values to reference variables in such a way that both $\phi_x$ and $\cfgnt{r}_f =
\cfgnt{r}_\mathit{null}$ hold). (ii)
$\theta_\mathit{new}$ is active when $\cfgnt{r}_f$ points to the
fresh location. This occurs when $\cfgnt{r}_f$ is not new or any other alias. (iii) each member of $\theta_\mathit{alias}$ restricts $\cfgnt{r}_f$ to
a particular alias in $\rho$, and at the same time, not alias any member of $\rho$
that was initialized earlier than the current choice (a later
initialization may indeed also alias the current choice in a valid
heap as order determines the number of aliasing choices available). And (iv) $\theta_\mathit{orig}$  implements conditional initialization to preserve the
original heap structure. This property is needed for the bisimulation. 
These sets are added into the heap on $\cfgnt{r}_f$ after the fields for $\cfgnt{l}_f$ are initialized to $\cfgnt{r}_\mathit{un}$.

\begin{figure*}[t]
\begin{center}
\begin{tabular}[c]{c|c|c|c}
\begin{tabular}[c]{c}
\scalebox{0.81}{\input{origHeap.pdf_t}} \\
\end{tabular} &
\begin{tabular}[c]{c}
\scalebox{0.81}{\input{summarizeXHeap.pdf_t}} \\
\end{tabular} &
\begin{tabular}[c]{c}
\scalebox{0.81}{\input{summarizeYHeap.pdf_t}} \\
\end{tabular} &
\begin{tabular}[c]{l}
$\rho := \{ (\cfgnt{r}_1^i\ \cfgnt{l}_1 \}$ \\
$\theta_\mathit{null} := \{ ( \cfgnt{r}_2^i = \cfgnt{r}_\mathit{null}\ \cfgnt{l}_\mathit{null}) \}$\\
$\theta_\mathit{new} := \{ ( \cfgnt{r}_2^i \neq \cfgnt{r}_\mathit{null} \wedge \cfgnt{r}_2^i \neq \cfgnt{r}_1^i\ \cfgnt{l}_2) \}$\\
$\theta_\mathit{alias} := \{ ( \cfgnt{r}_1^i \neq \cfgnt{r}_\mathit{null} \wedge \cfgnt{r}_2^i \neq \cfgnt{r}_\mathit{null} \wedge \cfgnt{r}_2^i = \cfgnt{r}_1^i\ \cfgnt{l}_1) \}$\\
$\theta_\mathit{orig} := \{ \}$ \\
$\phi_{\mathit{1a}} := \cfgnt{r}_1^i = \cfgnt{r}_\mathit{null} $ \\
$\phi_{\mathit{1b}} := \cfgnt{r}_1^i \neq \cfgnt{r}_\mathit{null} $  \\
$\phi_{\mathit{2a}} := \cfgnt{r}_2^i = \cfgnt{r}_\mathit{null}$ \\
$\phi_{\mathit{2b}} := \cfgnt{r}_2^i \neq \cfgnt{r}_\mathit{null} \wedge \cfgnt{r}_2^i \neq \cfgnt{r}_1^i$ \\
$\phi_{\mathit{2c}} :=  \cfgnt{r}_2^i \neq \cfgnt{r}_\mathit{null} \wedge \cfgnt{r}_2^i = \cfgnt{r}_1^i $ \\
\end{tabular} \\
(a) & (b) & (c) & (d) \\
\end{tabular}
\end{center}
\caption{An example that initializes $\lp\cfgt{this}\  \cfgt{\$}\ \cfgnt{x}\rp$ and $\lp\cfgt{this}\  \cfgt{\$}\ \cfgnt{y}\rp$. (a) Initial heap structure. (b) After $\lp\cfgt{this}\  \cfgt{\$}\ \cfgnt{x}\rp$ is initialized. (c) After $\lp\cfgt{this}\  \cfgt{\$}\ \cfgnt{y}\rp$ is initialized. (d) Sets in the summarize rule and constraints on the edges.}
\label{fig:initHeap}
\end{figure*}

We visualize the initialization process in~\figref{fig:initHeap}. The
graph in~\figref{fig:initHeap}(a) represents the initial heap. The
reference superscripts $s$ and $i$ indicate the partition containing the reference.
 In~\figref{fig:initHeap}, $\cfgnt{r}_0^s$ represents a stack
reference for the $\cfgt{this}$ variable which has two fields $x$
and $y$ of the same type. Note that when no constraint is specified for a location,
there is an implicit $\mathit{true}$ constraint. For example, $\cfgnt{r}_0^s$
points to $\cfgnt{l}_0$ on the constraint $\mathit{true}$. The fields $x$ and
$y$ point to the uninitialized reference $\cfgnt{r}_\mathit{un}$.

The graph in~\figref{fig:initHeap}(b) represents the summary heap
after the initialization of the $\lp\cfgt{this}\  \cfgt{\$}\ \cfgnt{x}\rp$ field while
the graph in~\figref{fig:initHeap}(c) represents the symbolic heap
summary after the initialization of the $\lp\cfgt{this}\  \cfgt{\$}\ \cfgnt{y}\rp$ field
following the initialization of $\lp\cfgt{this}\  \cfgt{\$}\ \cfgnt{x}\rp$. The list
in~\figref{fig:initHeap}(d) represents the various sets constructed in
the summarize rule during the initialization of $\lp\cfgt{this}\  \cfgt{\$}\ \cfgnt{y}\rp$. We
also define values of constraints used as labels in the graph.

Accessing field $x$ from $\cfgnt{l}_0$ creates a new input heap reference $\cfgnt{r}_1^i$ in the
summary heap shown in~\figref{fig:initHeap}(b). Reference $\cfgnt{r}_1^i$
points to the $\cfgnt{l}_\mathit{null}$ location under the constraint that
$\cfgnt{r}_1^i$ is null (constraint $\phi_{1a}$ in  \figref{fig:initHeap}(d)), reference
$\cfgnt{r}_1^i$ points to location $\cfgnt{l}_1$ under the constraint that $\cfgnt{r}_1^i$
is not $\mathit{null}$ (constraint $\phi_{1b}$), and fresh location $\cfgnt{l}_1$ of type $C$ 
(the type of field $\cfgnt{x}$), is added to the heap. There are no aliases because $\cfgnt{l}_1$ 
is the first location of type $C$ to be initialized.

Following $\lp\cfgt{this}\ \cfgt{\$}\ \cfgnt{x}\rp$,
$\lp\cfgt{this}\ \cfgt{\$}\ \cfgnt{y}\rp$ is initialized leading to
the summary heap in~\figref{fig:initHeap}(c). The set $\rho$
in~\figref{fig:initHeap}(d) contains the one potential aliases
$\cfgnt{l}_1$ from the previous initialization for
$\lp\cfgt{this}\ \cfgt{\$}\ \cfgnt{x}\rp$. It also shows the
$\theta$-sets for this initialization and the resulting edge
labels. Note that the ordering between aliases in
$\theta_\mathit{alias}$ is insignificant until a third field is
initialized and there are two potential choices.
